\Title

### 2022-09-28 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

Learning hetero-synaptic delays in a layer of spiking neurons for fast motion detection

@@

The response of a biological neuron depends on the precise timing of afferent spikes. This temporal aspect of the neuronal code is essential in understanding information processing in neurobiology and applies particularly well to the output of neuromorphic hardware such as event-based cameras. However, most artificial neuronal models do not take advantage of this minute temporal dimension. Inspired by this neuroscientific observation, we develop a model for the efficient detection of temporal spiking motifs based on a layer of neurons with hetero-synaptic delays. Indeed, the variety of synaptic delays on the dendritic tree allows to synchronize synaptic inputs as they reach the basal dendritic tree. We show this can be formalized as a time-invariant logistic regression which can be trained using labelled data. We apply this model to solve the specific computer vision problem of motion detection, and demonstrate its application to synthetic naturalistic videos transformed into event streams similar to the output of event-based cameras. In particular, we quantify how its accuracy can vary with the total computational load. This end-to-end event-driven computational brick could help improve the performance of future Spiking Neural Network (SNN) algorithms and their prospective use in neuromorphic chips.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, estimated at 5W at rest. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second to an average of 10000 synapses, but each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks, and in particular to achieve the same performance as observed in classical CNNs, but with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on a property of spike neurons inspired by neurobiology to progress in this direction.
%
\subsection{What can neuroscience bring to computer vision?}% and the hetero-synaptic delays in spiking neurons}
%%%-----------------------------------------------------------------
Let us first focus on some existing spiking neural network models. A first motivation for these models was to understand biological observations at different scales. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential and the role of different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network (Brunel). Several models have increased the scale of the system to include more neurons, such as the Virtual Brain (Sanz), which mixes spike neural networks with analog neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few have had an application in computer vision. Another class of approaches have designed spiking neural networks that aim to have applications in machine learning. One of these is the SpikeNet algorithm which uses a radical approach by encoding information using one spike per neuron (Delorme). Another type of spiking neural network attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm (Bohte) and is now widely used in methods that attempt to transfer performance from CNNs to SNNs, for example using a surrogate gradient approach. However, the performance of SNNs is still lower than that of analog networks and the question of the advantage of using spikes remains open.

@@

This paper explores the possibility that spiking motifs may be a useful representation used in neurobiology and that we may transsfer this to neuromorphic algorithms. Our spiking neural network model is defined as a simple layer of leaky integrate-and-fire neurons whose connections are defined by a matrix of synaptic weights. Here, we propose to extend our model to a layer of spiking neurons which include, in addition to synaptic weights, hetero-synaptic delays. In particular, one afferent may be connected with multiple delays and, crucially, we will explicitly use the delay in the computational process. As a consequence, this extends our previous model by encoding spiking motifs as 3D images. The objective in this model, by including the dimension of temporal delays, is to increase the representational capacities of the classifier. In the perspective of building energy-efficient algorithms, we will also titrate quantitatively the best trade-off between robustness and computation time when increasing the number of these hetero-synaptic delays. %

@@

In this work, we study the emergence of such spatiotemporal spike motifs when training a single layer of spike neurons on a supervised classification task (\seeFig{motion_task}). The paper is organized as follows. We develop an SNN-like method capable of learning hetero-synaptic delays to perform motion detection on an event-based synthetic dataset.   We will first detail the methodology by defining the core spike mechanism using hetero-synaptic delays. We will then formally define the computational task of motion detection. This will allow us to formalize the final spiking neuron used to learn the model and test its effectiveness.  In the results section, we will detail the emergent structure of the model solving the task and show in particular the similarities with neurobiological observations. We will also test the network with classical stimuli used in neuroscience, and in particular the frequency or contrast responses.  We will also study the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. This will allow us to show how such a model can provide an effective solution to the energy/accuracy trade-off.  Finally, we will conclude by highlighting the main contributions of this paper, while defining some limitations. This will open up perspectives for future SNNs.  As neuromorphic devices are, by design, good candidates for integrating computations over time, we highlight the fact that this event-driven algorithm is transferable to this type of hardware.

@@

Let us now formally define the hetero-synaptic model, as well as the task and how we will learn to solve it. First, we will define the model for the efficient detection of event-based motifs using a feed-forward layer of spiking neural network (SNN) with hetero-synaptic delays. Our SNN model will be tested on a computationally relevant task of motion detection in an event-based sensory input. By analogy with biological conditions, we will design a paradigm simulating eye movements where an input image is translated by a parameterized saccadic trajectory. Solving is, for example, useful in real-life situations to compensate for eye, head or body movements and to provide robust image categorization by stabilizing the image on the retina. The task is thus to guess the motion as accurately as possible (see \seeFig{motion_task}) and we will describe a supervised learning mechanism for weights and delays. To this end, we will extend the SNN model to be efficient for the resolution of the task by including convolutional kernels. We will finally describe its implementation on conventional computers. 

@@

\subsection{Detecting event-based motifs using spiking neurons with hetero-synaptic delays}% model as a Logistic Regression with a temporal convolution}

@@

	    \caption{Detecting event-based motifs using spiking neurons with hetero-synaptic delays. 

@@

This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of hetero-synaptic delays, i.e., the precise timing from synaptic activation to its integration into the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that hetero-synaptic description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrite defined by this synaptic set and notably by the respective delays, which will multiplex in time all events.

Formally, let's define a layer of hetero-synaptic spiking neurons $\postsynaddr \in \postsynaddrspace$ by first describing how each neuron connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the hetero-synaptic model. As a consequence, an event stream $\event_\presynaddrspace$ on the presynaptic address space is transformed into a new event stream $\event_\postsynaddrspace$ which is defined by 
$\{ (\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr, \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr \}_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} )
\}$. \note{formaliser le multiplexage}
When post-synaptic neurons are activated on their soma by this discharge motif the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner.

@@

\note{ TODO: show why this formula implements a polychrony detector }
From the perspective of simulating such event-based computations on standard chips, it is useful to transform this sparse representation into a dense representation. As such, we may first write any event-based input as the boolean matrix $A \in \{0, 1 \}^\presynaddrspace$. In this simplified model, we will consider that hetero-synaptic delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
Using this dense representation, the counting defined above becomes:
\begin{equation*}
\mathcal{C}^\postsynaddr(a,t)
= \sum_{\presynaddr, \synapticdelay_\ranksyn^\postsynaddr} \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn^\postsynaddr)
\end{equation*}
%
This shows that $\mathcal{C}^\postsynaddr$ is a temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}^\postsynaddr = \kernel^\postsynaddr \ast A$.
This well-known computation defines a time-invariant, differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.
%

@@

\subsubsection{Implementation using spiking neurons with hetero-synaptic delays}
%
The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%

\note{say that it is effortless in biology but difficult on conventional computers}
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
%
\subsubsection{Performance on synthetic data}
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.490\linewidth]{figures/figure_N_PGs.pdf}
    \includegraphics[width=0.490\linewidth]{figures/figure_N_PG_time.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with hetero-synaptic delays. 
	    {\bf (a)}~Accuracy of PG detection as a function of the number $M$ of kernels.
	    {\bf (b)}~Accuracy of PG detection as a function of the temporal depth $D$ of kernels.
	    	     }
    \label{fig:model_results}
\end{figure}
% 
This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the PGs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider PG kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items.

To quantify the efficiency of this operation, we generated $M=55$ synthetic PGs as random independent kernels over $128$ presynaptic inputs and $D=127$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $2.0$ occurrences each. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=55$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the potential of hetero-synaptic delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve it. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.

@@

Once our MLR is trained, we obtain 3D kernels corresponding to the weights associated to the hetero-synaptic delays of our layer of spiking neurons and which may be used for detection. We observed that the distribution of the kernels' weights is sparse, with most values near zero. As shown in the formalization of our event-based model, the computational cost of our model if implemented on a neuromorphic chip would be dominated by the number of spikes times the number of synapses. This scales with the number of nonzero synaptic weights. To  assess the robustness of the classification as a function of the computational load, we will prune the weights in $\{\synapse_\ranksyn\}_{\ranksyn \in [0,\Nsyn)}$ that are below a defined threshold.

In Figure~\ref{fig:accuracy}, we plot the classification accuracy as a function of the relative number of computations, or active weights, per decision for each neuron of the layer. As a comparison and to account for the gain in performance by using hetero-synaptic delays, we provide the accuracy obtained with a MLR model using 2D time surface (in red) as in~\citep{grimaldi_robust_2022}. This latter method is based on delays from the last recorded events and uses fewer computations (in our case $15\times15$) than the dense 3D kernels without any pruning ($15\times15\times8$). While less computations are needed, the classification performance obtained for the model using time surfaces is similar to our method using all the weights of the kernels.
By pruning weights, we observe that the evolution of accuracy as a function of the log percentage of active weights follows a sigmoid curve. Half-saturation level is reached at about $3.5\times 10^{-3}\%$ of active weights, corresponding to a total amount of $6$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained to its top performances when dividing the number of computations by a factor up to about $200$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, thus demonstrating the efficiency of the presented method.

>>>>

Learning heterogeneous delays in a layer of spiking neurons for fast motion detection

@@

The response of a biological neuron depends on the precise timing of afferent spikes. This temporal aspect of the neuronal code is essential in understanding information processing in neurobiology and applies particularly well to the output of neuromorphic hardware such as event-based cameras. However, most artificial neuronal models do not take advantage of this minute temporal dimension. Inspired by this neuroscientific observation, we develop a model for the efficient detection of temporal spiking motifs based on a layer of neurons with heterogeneous delays. Indeed, the variety of synaptic delays on the dendritic tree allows to synchronize synaptic inputs as they reach the basal dendritic tree. We show this can be formalized as a time-invariant logistic regression which can be trained using labelled data. We apply this model to solve the specific computer vision problem of motion detection, and demonstrate its application to synthetic naturalistic videos transformed into event streams similar to the output of event-based cameras. In particular, we quantify how its accuracy can vary with the total computational load. This end-to-end event-driven computational brick could help improve the performance of future Spiking Neural Network (SNN) algorithms and their prospective use in neuromorphic chips.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, estimated at 5W at rest. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second to an average of 10000 synapses, but each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks, and in particular to achieve the same performance as observed in classical CNNs, but with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on a property of spike neurons inspired by neurobiology to progress in this direction.
%
\subsection{What can neuroscience bring to computer vision?}% and the heterogeneous delays in spiking neurons}
%%%-----------------------------------------------------------------
Let us first focus on some existing spiking neural network models. A first motivation for these models was to understand biological observations at different scales. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential and the role of different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network (Brunel). Several models have increased the scale of the system to include more neurons, such as the Virtual Brain (Sanz), which mixes spike neural networks with analog neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few have had an application in computer vision. Another class of approaches have designed spiking neural networks that aim to have applications in machine learning. One of these is the SpikeNet algorithm which uses a radical approach by encoding information using one spike per neuron (Delorme). Another type of spiking neural network attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm (Bohte) and is now widely used in methods that attempt to transfer performance from CNNs to SNNs, for example using a surrogate gradient approach. However, the performance of SNNs is still lower than that of analog networks and the question of the advantage of using spikes remains open.

@@

This paper explores the possibility that spiking motifs may be a useful representation used in neurobiology and that we may transsfer this to neuromorphic algorithms. Our spiking neural network model is defined as a simple layer of leaky integrate-and-fire neurons whose connections are defined by a matrix of synaptic weights. Here, we propose to extend our model to a layer of spiking neurons which include, in addition to synaptic weights, heterogeneous delays. In particular, one afferent may be connected with multiple delays and, crucially, we will explicitly use the delay in the computational process. As a consequence, this extends our previous model by encoding spiking motifs as 3D images. The objective in this model, by including the dimension of temporal delays, is to increase the representational capacities of the classifier. In the perspective of building energy-efficient algorithms, we will also titrate quantitatively the best trade-off between robustness and computation time when increasing the number of these heterogeneous delays. %

@@

In this work, we study the emergence of such spatiotemporal spike motifs when training a single layer of spike neurons on a supervised classification task (\seeFig{motion_task}). The paper is organized as follows. We develop an SNN-like method capable of learning heterogeneous delays to perform motion detection on an event-based synthetic dataset.   We will first detail the methodology by defining the core spike mechanism using heterogeneous delays. We will then formally define the computational task of motion detection. This will allow us to formalize the final spiking neuron used to learn the model and test its effectiveness.  In the results section, we will detail the emergent structure of the model solving the task and show in particular the similarities with neurobiological observations. We will also test the network with classical stimuli used in neuroscience, and in particular the frequency or contrast responses.  We will also study the efficiency of the motion detection mechanism and in particular its resilience to synaptic weight pruning. This will allow us to show how such a model can provide an effective solution to the energy/accuracy trade-off.  Finally, we will conclude by highlighting the main contributions of this paper, while defining some limitations. This will open up perspectives for future SNNs.  As neuromorphic devices are, by design, good candidates for integrating computations over time, we highlight the fact that this event-driven algorithm is transferable to this type of hardware.

@@

Let us now formally define the heterogeneous model, as well as the task and how we will learn to solve it. First, we will define the model for the efficient detection of event-based motifs using a feed-forward layer of spiking neural network (SNN) with heterogeneous delays. Our SNN model will be tested on a computationally relevant task of motion detection in an event-based sensory input. By analogy with biological conditions, we will design a paradigm simulating eye movements where an input image is translated by a parameterized saccadic trajectory. Solving is, for example, useful in real-life situations to compensate for eye, head or body movements and to provide robust image categorization by stabilizing the image on the retina. The task is thus to guess the motion as accurately as possible (see \seeFig{motion_task}) and we will describe a supervised learning mechanism for weights and delays. To this end, we will extend the SNN model to be efficient for the resolution of the task by including convolutional kernels. We will finally describe its implementation on conventional computers. 

@@

\subsection{Detecting event-based motifs using spiking neurons with heterogeneous delays}% model as a Logistic Regression with a temporal convolution}

@@

	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 

@@

This raster plot is generated by spiking neurons which are defined on the one hand by the equations governing the evolution of its membrane potential dynamics on its soma and on the other hand by the characterization of the synaptic contacts on its dendritic tree, or on the sensory processes (such as the pixel on the sensor of a camera, or the photoreceptors in the retina). A classical characterization consists in detailing the synaptic weights of each synaptic contact on the dendritic tree, the so-called weight matrix. A more detailed description of these synaptic contacts reveals the added importance of heterogeneous delays, i.e., the precise timing from synaptic activation to its integration into the soma, and how this changes the network's dynamics~\citep{izhikevich_polychronization_2006}. In that heterogeneous description, we can parameterize each neuron by the set of tuples defining both the weight and the delay of each synaptic contact. As a consequence, a set of input presynaptic spikes $\event$ will be processed by the dendrite defined by this synaptic set and notably by the respective delays, which will multiplex in time all events.

Formally, let's define a layer of heterogeneous spiking neurons $\postsynaddr \in \postsynaddrspace$ by first describing how each neuron connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the heterogeneous model. As a consequence, an event stream $\event_\presynaddrspace$ on the presynaptic address space is transformed into a new event stream $\event_\postsynaddrspace$ which is defined by 
$\{ (\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr, \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr \}_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} )
\}$. \note{formaliser le multiplexage}
When post-synaptic neurons are activated on their soma by this discharge motif the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner.

@@

\note{ TODO: show why this formula implements a polychrony detector }
From the perspective of simulating such event-based computations on standard chips, it is useful to transform this sparse representation into a dense representation. As such, we may first write any event-based input as the boolean matrix $A \in \{0, 1 \}^\presynaddrspace$. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
Using this dense representation, the counting defined above becomes:
\begin{equation*}
\mathcal{C}^\postsynaddr(a,t)
= \sum_{\presynaddr, \synapticdelay_\ranksyn^\postsynaddr} \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn^\postsynaddr)
\end{equation*}
%
This shows that $\mathcal{C}^\postsynaddr$ is a temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}^\postsynaddr = \kernel^\postsynaddr \ast A$.
This well-known computation defines a time-invariant, differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.
%

@@

\subsubsection{Implementation using spiking neurons with heterogeneous delays}
%
The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%

\note{say that it is effortless in biology but difficult on conventional computers}
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the heterogeneous delays model. 
%
\subsubsection{Performance on synthetic data}
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.490\linewidth]{figures/figure_N_PGs.pdf}
    \includegraphics[width=0.490\linewidth]{figures/figure_N_PG_time.pdf}
	    \caption{Detecting event-based motifs using spiking neurons with heterogeneous delays. 
	    {\bf (a)}~Accuracy of PG detection as a function of the number $M$ of kernels.
	    {\bf (b)}~Accuracy of PG detection as a function of the temporal depth $D$ of kernels.
	    	     }
    \label{fig:model_results}
\end{figure}
% 
This generative model defined above allows to determine this inference model for guessing sources $B$ when observing a raster plot $A$. This assumes that we know the PGs as defined by the $W_b$ matrices. The underlying metric is the binary cross-entropy, as used in the logistic regression model. In particular, if we consider PG kernels with similar decreasing exponential time profile, we can prove that this is similar to the method of~\citet{berens_fast_2012}. In our specific case, the difference is that the regression is performed in both dendritic and delay space by extending the summation using a temporal convolution operator. Using this forward model, it possible to estimate the logit (inverse of a sigmoid) $\hat{B}(b, t)$ for the presence of a PG of address $b$ and at time $t$ by using the transpose convolution operator. It thus comes that when observing $A$, then one may infer $\hat{B} = A \ast W^T$ and select the most activated items.

To quantify the efficiency of this operation, we generated $M=55$ synthetic PGs as random independent kernels over $128$ presynaptic inputs and $D=127$ possible delays. We drew random independent instances of $B$ with a length of $T=1000$ time steps and with on average $2.0$ occurrences each. This allowed us to generate raster plots which we use to infer $\hat{B}$. We compute the accuracy as the rate of true positive detections (both for inferring the address and its exact timing) and observe on average $\approx 98\%$ correct detections. We further extended this result by showing how the accuracy would evolve as a function of the number of simultaneous PGs, while keeping the same frequency of occurrence. We show in Figure~\ref{fig:model_results}~(a) that the accuracy of finding the right PG is still above $80\%$ accuracy with more than $1364$ overlapping PGs. Moreover, we show in Figure~\ref{fig:model_results}~(b) that (with $M=55$ PGs fixed) the accuracy increases notably as the temporal depth $D$ of the PG kernel increased, demonstrating quantitatively the potential of heterogeneous delays. These results were obtained while assuming that we know $W$. However, this is in general not the case, for instance when observing the raster plot of a population of neurons. In the following, we will define a generic visual task and determine a learning algorithm to solve it. %Inspired by the k-means algorithm, it is possible to devise a self-supervised learning algorithm. Our preliminary results show that it is possible to retrieve PGs embedded in the data, yet that further analysis is necessary to improve the convergence of the algorithm. In particular, it seems promising to use a sparseness constraint in the inference mechanism such as to remove spurious correlations in the inference.

@@

Once our MLR is trained, we obtain 3D kernels corresponding to the weights associated to the heterogeneous delays of our layer of spiking neurons and which may be used for detection. We observed that the distribution of the kernels' weights is sparse, with most values near zero. As shown in the formalization of our event-based model, the computational cost of our model if implemented on a neuromorphic chip would be dominated by the number of spikes times the number of synapses. This scales with the number of nonzero synaptic weights. To  assess the robustness of the classification as a function of the computational load, we will prune the weights in $\{\synapse_\ranksyn\}_{\ranksyn \in [0,\Nsyn)}$ that are below a defined threshold.

In Figure~\ref{fig:accuracy}, we plot the classification accuracy as a function of the relative number of computations, or active weights, per decision for each neuron of the layer. As a comparison and to account for the gain in performance by using heterogeneous delays, we provide the accuracy obtained with a MLR model using 2D time surface (in red) as in~\citep{grimaldi_robust_2022}. This latter method is based on delays from the last recorded events and uses fewer computations (in our case $15\times15$) than the dense 3D kernels without any pruning ($15\times15\times8$). While less computations are needed, the classification performance obtained for the model using time surfaces is similar to our method using all the weights of the kernels.
By pruning weights, we observe that the evolution of accuracy as a function of the log percentage of active weights follows a sigmoid curve. Half-saturation level is reached at about $3.5\times 10^{-3}\%$ of active weights, corresponding to a total amount of $6$ computations per decision. Compared to the full kernels, the accuracy of our method is maintained to its top performances when dividing the number of computations by a factor up to about $200$. In this case, the number of computations is greatly reduced compared to~\citep{grimaldi_robust_2022}, thus demonstrating the efficiency of the presented method.
